Co-methylation review

### 2015-09-24 Rmarkdown/comethylation_review.Rmd
Closes #16

In their study of differential methylation in queen and worker honey bees (_Apis mellifera_), \citet{Lyko:2010dr} perform an autocorrelation analysis of CpG methylation. They report that the "correlation of methylation status of neighbouring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away". This analysis, however, is fundamentally flawed.

Figure \ref{fig:Lyko} reproduces the results for the co-methylation analysis. Note that the maximum value of the "CpG autocorrelation" is less than $0.015$ (panel (A) of Figure \ref{fig:Lyko}). I strongly suspect that the autocorrelation has been incorrectly computed. Specifically, what I believe has been computed is the autocorrelation of all positions in the genome, not just CpGs, where non-CpGs have been artificially assigned a $\beta$-value of zero. It does not make sense to include non-CpG positions, particularly non-cytosines that can never be methylated, in an analysis of CpG co-methylation. I do not have the data used in \citet{Lyko:2010dr}, but the following example supports my conclusion. This example uses the chromosome 22 data from the _FF_ sample in the _Lister_ dataset (see Section \ref{sec:Lister_dataset} for full details of this sample).

@@

@@

Figure \ref{fig:FF_chr21_acf_beta_star} is a plot of the autocorrelation of the augmented vector, $\bm{\beta^{\ast}}$ for the chromosome 22 data from the _FF_ sample. We see that it is remarkably similar to that of panel (a) in Figure \ref{fig:Lyko}; the correlation first increases then decreases and the maximum correlation is very small, close to zero.

@@

Furthermore, if we compute the Fourier transform of this autocorrelation we find a period of approximately 3 bp, identical to that reported by \citet{Lyko:2010dr} for honey bees (compare panel (c) of Figure \ref{fig:Lyko} and Figure \ref{fig:FF_chr21_acf_beta_star_fourier}). Like my earlier criticism of the co-methylation results of \citet{Lister:2009hy}, I believe the apparent $3$ bp period is more likely due to the spacing of CpGs in the genome than anything to do with their methylation state.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star_fourier.pdf}
\caption[Fourier analysis of incorrectly defined $\beta$-values]{Fourier transformation of the autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \textit{FF} sample. Fourier transformation computed and plotted using the \texttt{spectrum} function in R. The period is defined as 1 / frequency. The maximum frequency is approximately $0.33$, which corresponds to a period of approximately $3$ bp.}
\label{fig:FF_chr21_acf_beta_star_fourier}
\end{figure}

All of this casts serious doubt on the validity and interpretation of the co-methylation results in \citet{Lyko:2010dr}.

>>>>

In their study of differential methylation in queen and worker honey bees (_Apis mellifera_), \citet{Lyko:2010dr} perform an autocorrelation analysis of CpG methylation. They report that the "correlation of methylation status of neighbouring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away". This analysis, however, is potentially flawed.

Figure \ref{fig:Lyko} reproduces the results for the co-methylation analysis. Note that the maximum value of the "CpG autocorrelation" is less than $0.015$ (panel (A) of Figure \ref{fig:Lyko}). I suspect that the autocorrelation has been incorrectly computed. Specifically, what I believe has been computed is the autocorrelation of all positions in the genome, not just CpGs, where non-CpGs have been artificially assigned a $\beta$-value of zero. It does not make sense to include non-CpG positions, particularly non-cytosines that can never be methylated, in an analysis of CpG co-methylation.

@@

I do not have the data used in \citet{Lyko:2010dr}, but the following example supports my belief. This example uses the chromosome 22 data from the _FF_ sample in the _Lister_ dataset (see Section \ref{sec:Lister_dataset} for full details of this sample).

@@

Figure \ref{fig:FF_chr21_acf_beta_star} is a plot of the autocorrelation of the augmented vector, $\bm{\beta^{\ast}}$ for the chromosome 22 data from the _FF_ sample. We see that it is remarkably similar to that of panel (A) in Figure \ref{fig:Lyko}; the correlation first increases then decreases and the maximum autocorrelation is very small, close to zero.

@@

On the one hand, this suggests that this may be the cause of the negligible autocorrelations in Figure S7 of \citet{Lyko:2010dr}. On the other hand, supposing that the maximum autocorrelation of the $\beta$-values is truly less than $0.015$, then this suggests that co-methylation itself is negligible in honeybees, in contrast to results from other organisms.

The effort required to process the Lyko dataset, and the insufficient detail in the paper to facilitate the reproducibility of the analysis, mean that I decided not to invest further effort into exploring potential interpretations of the co-methylation analysis of \citet{Lyko:2010dr}.
